---
title: "Model distances between simulated populations"
output: html_document
date: "2023-03-31"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = "/Users/jhuddlesfredhutch.org/projects/flu-forecasting-delays")

library(dplyr)
library(readr)
library(ggplot2)
library(rethinking)
```

## Load and prepare data

Load distances between populations at a given current timepoint and a corresponding future timepoint.
The future timepoint corresponds to the forecast horizon (in months) used to make the forecast.

```{r}
distances <- read_csv("results/model_inputs.csv")
```

Standardize submission delay and forecast horizon values for use in models.
Both predictors use units of "months".

```{r}
distances$horizon_s <- (distances$horizon - mean(distances$horizon)) / sd(distances$horizon)
distances$delay_s <- (distances$delay - mean(distances$delay)) / sd(distances$delay)
```

## Run prior predictive checks

Determine priors for each model parameter such that the possible output is within realistic bounds.

```{r}
# Make combinations of horizons and delays.
horizons <- unique(distances$horizon)
delays <- unique(distances$delay)
horizons_delays <- expand.grid(delay=delays, horizon=horizons)

# Define total samples.
n_samples_per_input <- 5000
n_inputs <- length(horizons_delays$delay)
n_samples <- n_samples_per_input * n_inputs

# Make vectors of horizons and delays.
horizon <- rep(horizons_delays$horizon, n_samples_per_input)
delay <- rep(horizons_delays$delay, n_samples_per_input)
```

Sample from the appropriate distribution for each prior.

```{r}
# Sample scale values.
prior.scale <- rexp(n_samples, 1)

# Sample intercept and coefficients.
a <- rnorm(n_samples, 0, 0.075)
b_s <- rnorm(n_samples, 0, 0.075)
b_h <- rnorm(n_samples, 0, 0.075)

# Calculate the rate of the gamma distribution.
prior.mu <- exp(a + (b_s * delay) + (b_h * horizon))

# Sample distances to the future.
prior.distance <- rgamma2(n_samples, prior.mu, prior.scale)
```

Build a data frame of the prior results.

```{r}
priors <- data.frame(
  horizon=horizon,
  delay=delay,
  distance=prior.distance
)
```

Summarize prior distances by feature.

```{r}
priors %>% group_by(horizon, delay) %>% summarize(median=median(distance), mean=mean(distance))
```

Plot prior distributions.

```{r}
ggplot(priors, aes(as.factor(horizon), distance, colour=as.factor(delay))) + geom_boxplot() + theme_classic()
ggsave("prior-predictive-simulation.png")
```

## Fit a simple model

Fit a simple linear model with a single intercept to represent the average distance between populations across all samples when standardized delay and horizon values are zero.

```{r}
simple.model <- ulam(
  alist(
    distance ~ dgamma2(mu, scale),
    scale ~ dexp(1),
    log(mu) <- a + b_s * delay + b_h * horizon,
    a ~ dnorm(0, 0.075),
    b_s ~ dnorm(0, 0.075),
    b_h ~ dnorm(0, 0.075)
  ),
  data=distances,
  iter=10000,
  chains=4,
  cores=4,
  log_lik=TRUE
)
```

Summarize the model.

```{r}
precis(simple.model)
```

Plot the model coefficients and HPDI (?).

```{r}
plot(simple.model)
```

Most of the variation in model parameters appears in the "scale" parameter of the gamma distribution.

Just to see what happens, I tried plotting the relationship between different predictors and distance using the rethinking code examples from page 113.
In this example, I fixed the submission delay to the maximum value which is the "realistic" delay.

```{r}
horizon.seq <- unique(distances$horizon)
delay.seq <- rep(max(distances$delay), length(horizon.seq))
pred_dat <- list(horizon=horizon.seq, delay=delay.seq)
mu <- link(simple.model, data=pred_dat)
mu.mean <- apply(mu, 2, mean)
mu.PI <- apply(mu, 2, PI, prob=0.89)
sim.distance <- sim(simple.model, data=pred_dat)
distance.PI <- apply(sim.distance, 2, PI, prob=0.89)
```

Plot the results.

```{r}
sim.distance.df <- cbind(
  data.frame(
    horizon_start=3,
    distance_start=sim.distance[, 1]
  ),
  data.frame(
    horizon_end=12,
    distance_end=sim.distance[, 4]
  )
)
ggplot(sim.distance.df, aes(x=horizon_start, xend=horizon_end, y=distance_start, yend=distance_end)) + geom_segment(alpha=0.1) + theme_classic() + geom_point(aes(x=3, y=mu.mean[1], col="red")) + geom_point(aes(x=6, y=mu.mean[2], col="red")) + geom_point(aes(x=9, y=mu.mean[3], col="red")) + geom_point(aes(x=12, y=mu.mean[4], col="red")) + labs(x="Forecast horizon (months)", y="Distance to the future (AAs)") + theme(legend.position="none") + scale_x_continuous(breaks=horizon.seq)
ggsave("posterior-simulations-with-realistic-delay.png")
```

```{r}
mu.mean
```

```{r}
mu.PI
```

```{r}
distance.PI
```

Repeat the above analysis of the posterior with samples extracted from posterior directly.

```{r}
post.samples <- extract.samples(simple.model)
fixed.delay <- 0
sim.distance <- sapply(horizon.seq, function(horizon)
  rgamma2(
    length(post.samples$a),
    exp(post.samples$a + post.samples$b_s * fixed.delay + post.samples$b_h * horizon),
    post.samples$scale
  )
)
mu.link <- function(horizon) exp(post.samples$a + post.samples$b_s * fixed.delay + post.samples$b_h * horizon)
mu <- sapply(horizon.seq, mu.link)
mu.mean <- apply(mu, 2 ,mean)
mu.HPDI <- apply(mu, 2, HPDI, prob=0.89)
distance.PI <- apply(sim.distance, 2, PI, prob=0.89)
```

Plot results.

```{r}
plot(distance ~ horizon, distances[distances$delay == fixed.delay,], col=col.alpha(rangi2, 0.5))
lines(horizon.seq, mu.mean)
shade(mu.HPDI, horizon.seq)
shade(distance.PI, horizon.seq)
```

Plot the distribution of average distances to the present when all data are available.

```{r}
average.distance.to.present <- exp(post.samples$a)
#average.distance.to.present <- rgamma2(length(post.samples$a), exp(post.samples$a), post.samples$scale)
mean.average.distance.to.present <- mean(average.distance.to.present)
HPDI.average.distance.to.present <- HPDI(average.distance.to.present)
hist(average.distance.to.present, xlab="average distance to present without delay (AAs)")
abline(v=mean.average.distance.to.present, col="red", lw=2)
abline(v=HPDI.average.distance.to.present[1], col="red", lw=2, lty="dashed")
abline(v=HPDI.average.distance.to.present[2], col="red", lw=2, lty="dashed")
```

Mean of the average distance to the present without delay:

```{r}
mean.average.distance.to.present
```

HPDI of the average distance to present without delay:

```{r}
HPDI(average.distance.to.present)
```

Summarize the effect of the forecast horizon on distance to the future.
This is the additional distance to the future on top of the average distance to the present represented by the intercept measured in AAs per year.
We expect this value to be effectively the substitution rate in AAs per year (~3 AAs / year).

```{r}
additional.distance.to.future <- (exp(post.samples$a + post.samples$b_h * 12) - exp(post.samples$a))
mean.additional.distance.to.future <- mean(additional.distance.to.future)
HPDI.additional.distance.to.future <- HPDI(additional.distance.to.future)
hist(additional.distance.to.future, xlab="rate of distance to the future (AAs/year)")
abline(v=mean.additional.distance.to.future, col="red", lw=2)
abline(v=HPDI.additional.distance.to.future[1], col="red", lw=2, lty="dashed")
abline(v=HPDI.additional.distance.to.future[2], col="red", lw=2, lty="dashed")
```

Mean additional distance to the future (AAs per year):

```{r}
mean.additional.distance.to.future
```

HPDI additional distance to the future (AAs per year):

```{r}
HPDI.additional.distance.to.future
```

Next, I summarized the effect of one unit change in horizon (1 month) to enable comparison with the effect of one unit of change in submission delay.

```{r}
additional.distance.to.future.per.month <- (exp(post.samples$a + post.samples$b_h) - exp(post.samples$a))
mean.additional.distance.to.future.per.month <- mean(additional.distance.to.future.per.month)
HPDI.additional.distance.to.future.per.month <- HPDI(additional.distance.to.future.per.month)
hist(additional.distance.to.future.per.month, xlab="rate of distance to the future (AAs/month)")
abline(v=mean.additional.distance.to.future.per.month, col="red", lw=2)
abline(v=HPDI.additional.distance.to.future.per.month[1], col="red", lw=2, lty="dashed")
abline(v=HPDI.additional.distance.to.future.per.month[2], col="red", lw=2, lty="dashed")
```

```{r}
mean.additional.distance.to.future.per.month
```

HPDI additional distance to the future (AAs per year):

```{r}
HPDI.additional.distance.to.future.per.month
```

Summarize the effect of submission delays on distance to the present.
This is the additional distance to the present on top of the average distance to the present represented by the intercept measured in AAs per year.

```{r}
additional.distance.to.present <- (exp(post.samples$a + post.samples$b_s) - exp(post.samples$a))
mean.additional.distance.to.present <- mean(additional.distance.to.present)
HPDI.additional.distance.to.present <- HPDI(additional.distance.to.present)
hist(additional.distance.to.present, xlab="rate of distance to the present (AAs/month)")
abline(v=mean.additional.distance.to.present, col="red", lw=2)
abline(v=HPDI.additional.distance.to.present[1], col="red", lw=2, lty="dashed")
abline(v=HPDI.additional.distance.to.present[2], col="red", lw=2, lty="dashed")
```

Mean additional distance to the present (AAs per year):

```{r}
mean.additional.distance.to.present
```

HPDI additional distance to the present (AAs per year):

```{r}
HPDI.additional.distance.to.present
```

We might consider three possible realistic scenarios for future practice of influenza genomic surveillance and vaccine development:

1. A change in vaccine development reduces the required forecast horizon from 12 months to 6 months (e.g., through adoption of mRNA-based vaccines).
2. A change in genomic surveillance capacity and policy reduces the average submission delay of genomes to GISAID from 3 months to 1 month.
3. Both changes to vaccine development and genomic surveillance occur at once.

In the first scenario, we fix the submission delay to 3 months on average and calculate the difference between distances to the future for forecast horizon values of 12 and 6 months.

```{r}
fixed.delay <- 3
scenario.1.horizons <- c(6, 12)

scenario.1.distances <- sapply(scenario.1.horizons, function(horizon)
  rgamma2(
    length(post.samples$a),
    exp(post.samples$a + post.samples$b_s * fixed.delay + post.samples$b_h * horizon),
    post.samples$scale
  )
)

# Reduction in distance to the future expected by reducing forecast horizon from 12 to 6 months.
scenario.1.improvement <- scenario.1.distances[, 2] - scenario.1.distances[, 1]
mean.scenario.1.improvement <- mean(scenario.1.improvement)
HPDI.scenario.1.improvement <- HPDI(scenario.1.improvement)
hist(scenario.1.improvement, xlab="Reduction in distance to future for 12- to 6-month horizon (AAs)")
abline(v=mean.scenario.1.improvement, col="red", lwd=2)
abline(v=HPDI.scenario.1.improvement[1], col="red", lwd=2, lty="dashed")
abline(v=HPDI.scenario.1.improvement[2], col="red", lwd=2, lty="dashed")
```

Mean improvement:

```{r}
mean.scenario.1.improvement
```

89% HPDI improvement:

```{r}
HPDI.scenario.1.improvement
```

In the second scenario, we fix the forecast horizon to 12 months and calculate the difference between distances to the future for average submission delays of 3 and 1 months.

```{r}
fixed.horizon <- 12
scenario.2.delays <- c(1, 3)

scenario.2.distances <- sapply(scenario.2.delays, function(delay)
  rgamma2(
    length(post.samples$a),
    exp(post.samples$a + post.samples$b_s * delay + post.samples$b_h * fixed.horizon),
    post.samples$scale
  )
)

# Reduction in distance to the future expected by reducing average submission delay from 3 months to 1 month.
scenario.2.improvement <- scenario.2.distances[, 2] - scenario.2.distances[, 1]
mean.scenario.2.improvement <- mean(scenario.2.improvement)
HPDI.scenario.2.improvement <- HPDI(scenario.2.improvement)
hist(scenario.2.improvement, xlab="Reduction in distance to future for 3- to 1-month delay (AAs)")
abline(v=mean.scenario.2.improvement, col="red", lwd=2)
abline(v=HPDI.scenario.2.improvement[1], col="red", lwd=2, lty="dashed")
abline(v=HPDI.scenario.2.improvement[2], col="red", lwd=2, lty="dashed")
```

Mean improvement:

```{r}
mean.scenario.2.improvement
```

89% HPDI improvement:

```{r}
HPDI.scenario.2.improvement
```

In the last scenario, we calculate the difference between distances to the future for the status quo (horizon=12 months, delay=3 months) and the scenario outcome (horizon=6 months, delay=1 month).

```{r}
scenario.3.params <- rbind(
  list(horizon=6, delay=1),
  list(horizon=12, delay=3)
)

scenario.3.link <- function(horizon, delay) {
  rgamma2(
    length(post.samples$a),
    exp(post.samples$a + post.samples$b_s * delay + post.samples$b_h * horizon),
    post.samples$scale
  )
}

scenario.3a.distances <- scenario.3.link(scenario.3.params[1, ]$horizon, scenario.3.params[1, ]$delay)
scenario.3b.distances <- scenario.3.link(scenario.3.params[2, ]$horizon, scenario.3.params[2, ]$delay)

# Reduction in distance to the future expected by reducing forecast horizon from 12 to 6 months
# and average submission delay from 3 months to 1 month.
scenario.3.improvement <- scenario.3b.distances - scenario.3a.distances
mean.scenario.3.improvement <- mean(scenario.3.improvement)
HPDI.scenario.3.improvement <- HPDI(scenario.3.improvement)
hist(scenario.3.improvement, xlab="Reduction in distance to future for 6-month horizon, 1-month delay (AAs)")
abline(v=mean.scenario.3.improvement, col="red", lwd=2)
abline(v=HPDI.scenario.3.improvement[1], col="red", lwd=2, lty="dashed")
abline(v=HPDI.scenario.3.improvement[2], col="red", lwd=2, lty="dashed")
```

Mean improvement:

```{r}
mean.scenario.3.improvement
```

89% HPDI improvement:

```{r}
HPDI.scenario.3.improvement
```

Combine the data from each scenario to produce a single summary figure.

```{r}
improvements <- rbind(
  data.frame(id=1, scenario="Improved vaccine development (6-month horizon, 3-month delay)", improvement=scenario.1.improvement),
  data.frame(id=2, scenario="Improved genomic surveillance (12-month horizon, 1-month delay)", improvement=scenario.2.improvement),
  data.frame(id=3, scenario="Improved vaccine and surveillance (6-month horizon, 1-month delay)", improvement=scenario.3.improvement)
)
improvements$title <- reorder(improvements$scenario, improvements$id)
```

Combine summary statistics from each scenario.

```{r}
improvements.stats <- data.frame(
  id=c(1, 2, 3),
  scenario=c(
    "Improved vaccine development (6-month horizon, 3-month delay)",
    "Improved genomic surveillance (12-month horizon, 1-month delay)",
    "Improved vaccine and surveillance (6-month horizon, 1-month delay)"
  ),
  mean=c(mean.scenario.1.improvement, mean.scenario.2.improvement, mean.scenario.3.improvement),
  lower_hpdi=c(HPDI.scenario.1.improvement[1], HPDI.scenario.2.improvement[1], HPDI.scenario.3.improvement[1]),
  upper_hpdi=c(HPDI.scenario.1.improvement[2], HPDI.scenario.2.improvement[2], HPDI.scenario.3.improvement[2])
)
improvements.stats$title <- reorder(improvements.stats$scenario, improvements.stats$id)
```

Plot improvements.

```{r}
ggplot(improvements, aes(x=improvement)) + geom_vline(aes(xintercept=0), colour="#999999") + geom_histogram(breaks=seq(-10, 15, 1)) + facet_wrap(vars(title), nrow=3, labeller=labeller(id=function(id) { })) + theme_classic() + labs(x="Reduction in distance to the future (AAs)", y="Number of posterior samples") + theme(strip.background = element_blank(), title=element_text(size=12), strip.text=element_text(size=12)) + geom_vline(data=improvements.stats, aes(xintercept=mean, colour="red"), show.legend=FALSE) + geom_vline(data=improvements.stats, aes(xintercept=lower_hpdi, colour="red"), linetype="dashed", show.legend=FALSE) + geom_vline(data=improvements.stats, aes(xintercept=upper_hpdi, colour="red"), linetype="dashed", show.legend=FALSE)
ggsave("reduction-in-distance-to-future-by-scenario-for-simulated-populations.png", dpi=300)
```
```{r}
improvements.stats
```

I also wanted to know how overly optimistic our current forecasts are when the 3-month average submission delay is ignored.
Here, I fix the horizon to 12 and compare distances between 0 and 3 month delays.

```{r}
fixed.horizon <- 12
scenario.4.delays <- c(0, 3)

scenario.4.distances <- sapply(scenario.4.delays, function(delay)
  rgamma2(
    length(post.samples$a),
    exp(post.samples$a + post.samples$b_s * delay + post.samples$b_h * fixed.horizon),
    post.samples$scale
  )
)

# Reduction in distance to the future expected by reducing average submission delay from 3 months to 0 month.
scenario.4.improvement <- scenario.4.distances[, 2] - scenario.4.distances[, 1]
mean.scenario.4.improvement <- mean(scenario.4.improvement)
HPDI.scenario.4.improvement <- HPDI(scenario.4.improvement)
hist(scenario.4.improvement, xlab="Reduction in distance to future for 3- to 0-month delay (AAs)")
abline(v=mean.scenario.4.improvement, col="red", lwd=2)
abline(v=HPDI.scenario.4.improvement[1], col="red", lwd=2, lty="dashed")
abline(v=HPDI.scenario.4.improvement[2], col="red", lwd=2, lty="dashed")
```

Mean improvement:

```{r}
mean.scenario.4.improvement
```

89% HPDI improvement:

```{r}
HPDI.scenario.4.improvement
```

Next, I wanted to see the distributions of the predictor coefficients at a higher resolution than the precis plot above.

```{r}
prior.samples <- extract.prior(simple.model, n=10000)
post.samples <- extract.samples(simple.model)
```

Coefficient for submission delay.

```{r}
plot(density(post.samples$b_s), xlim=c(-0.1, 0.16))
lines(density(prior.samples$b_s), col="red")
abline(v=mean(post.samples$b_s), lty=2)
```

Coefficient for forecast horizon.

```{r}
plot(density(post.samples$b_h), xlim=c(-0.1, 0.18))
lines(density(prior.samples$b_h), col="red")
abline(v=mean(post.samples$b_h), lty=2)
```

Intercept representing average distance to the future across timepoints.

```{r}
plot(density(post.samples$a), xlim=c(-0.125, 1.5))
lines(density(prior.samples$a), col="red")
abline(v=mean(post.samples$a), lty=2)
```

Scale of gamma distribution representing the variance around the central rate.

```{r}
plot(density(post.samples$scale), xlim=c(-0.125, 1.75))
lines(density(prior.samples$scale), col="red")
abline(v=mean(post.samples$scale), lty=2)
```

### Posterior predictive checks

```{r}
horizons <- unique(distances$horizon)
delays <- unique(distances$delay)
horizons_delays <- expand.grid(delay=delays, horizon=horizons)
post <- sim(simple.model, horizons_delays)
```

Map treatments to posterior samples.
TODO: Do this a better way.

```{r}
all.post <- rbind(
  data.frame(horizon=3, delay=0, distance=post[,1]),
  data.frame(horizon=3, delay=1, distance=post[,2]),
  data.frame(horizon=3, delay=3, distance=post[,3]),
  data.frame(horizon=6, delay=0, distance=post[,4]),
  data.frame(horizon=6, delay=1, distance=post[,5]),
  data.frame(horizon=6, delay=3, distance=post[,6]),
  data.frame(horizon=9, delay=0, distance=post[,7]),
  data.frame(horizon=9, delay=1, distance=post[,8]),
  data.frame(horizon=9, delay=3, distance=post[,9]),
  data.frame(horizon=12, delay=0, distance=post[,10]),
  data.frame(horizon=12, delay=1, distance=post[,11]),
  data.frame(horizon=12, delay=3, distance=post[,12])
)
```

Plot the posterior distributions by treatment.

```{r}
ggplot(all.post, aes(as.factor(horizon), distance, colour=as.factor(delay))) + geom_boxplot() + theme_classic() + labs(x="Forecast horizon (months)", y="Distance to the future (AAs)", colour="Mean submission delay (months)") + theme(legend.position=c(0.25, 0.8))
ggsave("posterior-predictive-simulations-across-treatments.png")
```

Plot observed distances.

```{r}
ggplot(distances, aes(as.factor(horizon), distance, colour=as.factor(delay))) + geom_boxplot() + theme_classic() + labs(x="Forecast horizon (months)", y="Distance to the future (AAs)", colour="Mean submission delay (months)") + theme(legend.position=c(0.25, 0.8))
ggsave("observed-distances-for-simulations-across-treatments.png")
```

Summarize distances from observed data.

```{r, rows.print=20}
observed.median.distances <- distances %>% group_by(horizon, delay) %>% summarize(median=median(distance))
observed.median.distances
```

Then, summarize the corresponding distances from posterior predictive simulations.

```{r, rows.print=20}
post.median.distances <- all.post %>% group_by(horizon, delay) %>% summarize(median=median(distance))
post.median.distances
```

```{r}
observed.median.distances$type <- "observed"
post.median.distances$type <- "posterior"
all.median.distances <- rbind(observed.median.distances, post.median.distances)
```

Plot observed and median distances by factor.

```{r}
ggplot(all.median.distances, aes(as.factor(horizon), median, colour=as.factor(delay))) + geom_point(aes(shape=type)) + theme_classic() + labs(x="Forecast horizon (months)", y="Distance to the future (AAs)", colour="Mean submission delay (months)", shape="Type") + theme(legend.position=c(0.25, 0.7))
ggsave("observed-and-posterior-median-distances-for-simulations-across-treatments.png")
```

## Fit a multilevel model

Assigns an intercept to each timepoint with a shared mean, allowing the model to account for timepoint-specific variation.

```{r, include=FALSE}
full.model <- ulam(
  alist(
    distance ~ dgamma2(mu, scale),
    scale ~ dexp(1),
    log(mu) <- a + a_timepoint[t] + b_s * delay + b_h * horizon,
    a ~ dnorm(0, 0.075),
    a_timepoint[t] ~ dnorm(0, sigma),
    sigma ~ dexp(1),
    b_s ~ dnorm(0, 0.075),
    b_h ~ dnorm(0, 0.075)
  ),
  data=distances,
  iter=10000,
  chains=4,
  cores=4,
  log_lik=TRUE
)
```

Summarize top-level model.

```{r}
precis(full.model)
```

Plot primary model parameters.

```{r}
plot(full.model)
```

```{r}
full.post.samples <- extract.samples(full.model)
```

```{r}
dens(full.post.samples$sigma, xlim=c(-0.25, 3), ylim=c(0, 25))
dens(full.post.samples$a, add=TRUE)
```

```{r}
dens(full.post.samples$b_s, col="blue", ylim=c(0, 300), xlim=c(0, 0.1))
dens(full.post.samples$b_h, add=TRUE, col="orange")
```

Summarize complete model.

```{r}
precis(full.model, depth=2)
```

Plot all model parameters.

```{r fig.width=6, fig.height=10}
plot(full.model, depth=2)
```

Plot variation of timepoint intercepts in temporal order.

```{r}
plot(coef(full.model)[2:97], type="l")
```

```{r}
hist(coef(full.model)[2:97])
```

```{r}
compare(simple.model, full.model)
```

### Posterior predictive checks

```{r}
post <- sim(full.model, horizons_delays)
```

Map treatments to posterior samples.
TODO: Do this a better way.

```{r}
all.post <- rbind(
  data.frame(horizon=6, delay=0, distance=post[,1]),
  data.frame(horizon=6, delay=1, distance=post[,2]),
  data.frame(horizon=6, delay=3, distance=post[,3]),
  data.frame(horizon=12, delay=0, distance=post[,4]),
  data.frame(horizon=12, delay=1, distance=post[,5]),
  data.frame(horizon=12, delay=3, distance=post[,6])
)
```

Plot the posterior distributions by treatment.

```{r}
ggplot(all.post, aes(as.factor(horizon), distance, colour=as.factor(delay))) + geom_boxplot() + theme_classic() + labs(x="Forecast horizon (months)", y="Distance to the future (AAs)", colour="Mean submission delay (months)") + theme(legend.position=c(0.25, 0.8))
ggsave("posterior-predictive-simulations-across-treatments.png")
```

Summarize distances from observed data.

```{r}
distances %>% group_by(horizon, delay) %>% summarize(median=median(distance))
```
Then, summarize the corresponding distances from posterior predictive simulations.

```{r}
all.post %>% group_by(horizon, delay) %>% summarize(median=median(distance))
```
The simple model systematically underestimates the distance to the future for 6-month forecasts and overestimates the distance for 12-month forecasts.
I wonder whether allowing each timepoint to have its own intercept in a multilevel model would better account for this issue.

## Fit a multilevel model with a Gaussian process

The high variance of the variance parameter for time-specific intercepts in the full model and relatively high degree of overfitting by that model compared to the simple model suggested that something we wrong with the full model.
Reading more about these models, I realized that my varying intercept variable of time is a continuous variable with correlated values.
As such, I should probably use a Gaussian process (GP) to represent the covariance between timepoints.

Read in the pairwise distances between timepoints.

```{r}
distances.between.years <- as.matrix(read.csv("results/distance_in_years.csv"))
```

Fit the model with a GP.

```{r}
gp.model <- ulam(
  alist(
    distance ~ dgamma2(mu, scale),
    scale ~ dexp(1),
    log(mu) <- a + a_timepoint[t] + b_s * delay + b_h * horizon,
    vector[96]: a_timepoint ~ multi_normal( 0 , SIGMAT ),
    matrix[96,96]: SIGMAT <- cov_GPL2( Dmat , etasq , rhosq , sigmasq ),
    etasq ~ dexp(1),
    rhosq ~ dexp(1),
    sigmasq ~ dexp(1),
    a ~ dnorm(0, 0.075),
    b_s ~ dnorm(0, 0.075),
    b_h ~ dnorm(0, 0.075)
  ),
  data=list(
    t=distances$t,
    delay=distances$delay,
    horizon=distances$horizon,
    Dmat=distances.between.years,
    distance=distances$distance
  ),
  iter=10000,
  chains=4,
  cores=4,
  log_lik=TRUE
)
```

```{r}
precis(gp.model)
```

```{r}
plot(precis(gp.model))
```

```{r}
compare(simple.model, full.model, gp.model)
```

